\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-02-03 analysis/paper/casebase-manuscript.Rmd
removed CS4

options(kableExtra.latex.load_packages = FALSE)
library(casebase)
library(colorspace)
library(cowplot)
library(data.table)
library(glmnet)
library(kableExtra)
library(magrittr)
library(pracma)
library(flexsurv) # for fitting gompertz AFT
library(prodlim)
library(riskRegression)
library(splines)
library(survival)
library(tibble)
library(visreg)
library(lubridate)
library(tidyverse)
library(scales)
options(digits = 2)
# set the seed for reproducible output
set.seed(1234)

# Turn on/off sections for faster compilation
eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- TRUE
eval_cs4 <- TRUE
eval_colophon <- FALSE # Only for draft
echo_plot_code <- TRUE
# Define theme for plots
paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
# Define colour palette for CS 3
q4 <- qualitative_hcl(4, palette = "Dark 3")
names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
```

@@

# Case study 4---Stanford Heart Transplant Data

@@

@@

Since the exposure is time-dependent, we need to manually define the exposure variable *after* case-base sampling and *before* fitting the hazard function. For this reason, we use the `sampleCaseBase` function directly.

```{r stanford_cb, eval = eval_cs4, echo = TRUE}
cb_data <- sampleCaseBase(jasa,
  time = "futime",
  event = "fustat", ratio = 100
)
```

Next, we compute the number of days from acceptance into the program to transplant, and we use this variable to determine whether each population-moment is exposed or not.

```{r stanford_exp, eval = eval_cs4, dependson="stanford_cb", echo = TRUE}
# Define exposure variable
cb_data <- mutate(cb_data,
  txtime = time_length(accept.dt %--% tx.date,
                       unit = "days"
  ),
  exposure = case_when(
    is.na(txtime) ~ 0L, # No transplant
    txtime > futime ~ 0L,
    txtime <= futime ~ 1L
  )
)
```

Finally, we can fit the hazard using various linear predictors.

```{r stanford_fit, eval = eval_cs4, warning = FALSE, echo=TRUE, dependson="stanford_exp"}
# Fit several models
fit1 <- fitSmoothHazard(fustat ~ exposure,
  data = cb_data, time = "futime"
)
fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
  data = cb_data, time = "futime"
)
fit3 <- fitSmoothHazard(fustat ~ exposure * futime,
  data = cb_data, time = "futime"
)
```

Note that the third model includes an interaction term between exposure and follow-up time. In other words, this model no longer exhibits proportional hazards. The evidence of non-proportionality of hazards in the Stanford Heart Transplant data has been widely discussed [@arjas1988graphical].

We can then compare the goodness of fit of these three models using the Akaike Information Criterion (AIC).

@@

As we can see, the best fit is the third model. By visualizing the hazard functions for both exposed and unexposed individuals, we can more clearly see how the hazards are no longer proportional. We can easily obtain a plot of the hazards by using the `plot.singleEventCB` method:

```{r stanford-hazard, eval = eval_cs4, echo=TRUE, dependson="stanford_fit", message = FALSE, results="hide", fig.cap="\\label{fig:cs4hazardPlot} Plot of the hazard function for exposed and unexposed individuals."}
plot(fit3, hazard.params = list(xvar = "futime",
                                by = "exposure",
                                alpha = 0.05,
                                ylab = "Hazard",
                                data = cb_data))
```

Looking at Figure \ref{fig:cs4hazardPlot}, the non-proportionality seems to be more pronounced at the beginning of follow-up than the end. Finally, in Figure \ref{fig:cs4CIF}, we turn these estimates of the hazard function into estimates of the cumulative incidence functions.

```{r stanford-risk, eval = eval_cs4, echo = TRUE, dependson="stanford_fit", fig.cap="\\label{fig:cs4CIF} Plot of the cumulative incidence function for exposed and unexposed individuals."}
# Compute absolute risk curves
newdata <- data.frame(exposure = c(0, 1))
absrisk <- absoluteRisk(fit3,
  newdata = newdata,
  time = seq(0, 425, length.out = 100)
)

@@

As we can see in the above case-study, the \pkg{casebase} package can also be used to model time-varying exposures and non-proportional hazard functions.

>>>>

options(kableExtra.latex.load_packages = FALSE)
library(casebase)
library(colorspace)
library(cowplot)
library(data.table)
library(glmnet)
library(kableExtra)
library(magrittr)
library(pracma)
library(flexsurv) 
library(prodlim)
library(riskRegression)
library(splines)
library(survival)
library(tibble)
library(visreg)
library(lubridate)
library(tidyverse)
library(scales)
options(digits = 2)
# set the seed for reproducible output
set.seed(1234)

# Turn on/off sections for faster compilation
eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- TRUE
eval_cs4 <- FALSE
eval_colophon <- FALSE # Only for draft
echo_plot_code <- TRUE
# Define theme for plots
paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
# Define colour palette for CS 3
q4 <- qualitative_hcl(4, palette = "Dark 3")
names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
```

@@

<!--
# Case study 4---Stanford Heart Transplant Data

@@

-->

@@

<!--
Since the exposure is time-dependent, we need to manually define the exposure variable *after* case-base sampling and *before* fitting the hazard function. For this reason, we use the `sampleCaseBase` function directly.
-->

```{r stanford_cb, eval = eval_cs4}
cb_data <- sampleCaseBase(jasa,
  time = "futime",
  event = "fustat", ratio = 100
)
```

<!--
Next, we compute the number of days from acceptance into the program to transplant, and we use this variable to determine whether each population-moment is exposed or not.

-->

```{r stanford_exp, eval = eval_cs4, dependson="stanford_cb"}
# Define exposure variable
cb_data <- mutate(cb_data,
  txtime = time_length(accept.dt %--% tx.date,
                       unit = "days"
  ),
  exposure = case_when(
    is.na(txtime) ~ 0L, # No transplant
    txtime > futime ~ 0L,
    txtime <= futime ~ 1L
  )
)
```

<!--
Finally, we can fit the hazard using various linear predictors.
-->

```{r stanford_fit, eval = eval_cs4, warning = FALSE, dependson="stanford_exp"}
# Fit several models
fit1 <- fitSmoothHazard(fustat ~ exposure,
  data = cb_data, time = "futime"
)
fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
  data = cb_data, time = "futime"
)
fit3 <- fitSmoothHazard(fustat ~ exposure * futime,
  data = cb_data, time = "futime"
)
```

<!--
Note that the third model includes an interaction term between exposure and follow-up time. In other words, this model no longer exhibits proportional hazards. The evidence of non-proportionality of hazards in the Stanford Heart Transplant data has been widely discussed [@arjas1988graphical].

We can then compare the goodness of fit of these three models using the Akaike Information Criterion (AIC).
-->

@@

<!--
As we can see, the best fit is the third model. By visualizing the hazard functions for both exposed and unexposed individuals, we can more clearly see how the hazards are no longer proportional. We can easily obtain a plot of the hazards by using the `plot.singleEventCB` method:
-->

```{r stanford-hazard, eval = eval_cs4, dependson="stanford_fit", message = FALSE, results="hide", fig.cap="\\label{fig:cs4hazardPlot} Plot of the hazard function for exposed and unexposed individuals."}
plot(fit3, hazard.params = list(xvar = "futime",
                                by = "exposure",
                                alpha = 0.05,
                                ylab = "Hazard",
                                data = cb_data))
```

<!--
Looking at Figure \ref{fig:cs4hazardPlot}, the non-proportionality seems to be more pronounced at the beginning of follow-up than the end. Finally, in Figure \ref{fig:cs4CIF}, we turn these estimates of the hazard function into estimates of the cumulative incidence functions.
-->

```{r stanford-risk, eval = eval_cs4, dependson="stanford_fit", fig.cap="\\label{fig:cs4CIF} Plot of the cumulative incidence function for exposed and unexposed individuals."}
# Compute absolute risk curves
newdata <- data.frame(exposure = c(0, 1))
absrisk <- absoluteRisk(fit3,
  newdata = newdata,
  time = seq(0, 425, length.out = 100)
)

@@

<!--
As we can see in the above case-study, the \pkg{casebase} package can also be used to model time-varying exposures and non-proportional hazard functions. 
-->
